---
title: "STAT-370-PROJ-EDA" 
author: "William Quinn" 
format: html
self-contained: true
editor: visual
---

# Libraries {.appendix}

```{r}
library(RCurl)
library(ggplot2)
library(janitor)
library(dplyr)
library(ggsci)
theme_set(theme_classic())
```

# Importing {.appendix}

```{r}
EDA_GIT_LINK <- getURL("https://raw.githubusercontent.com/Cgrossman23/STAT-370-PROJ/refs/heads/main/cleaned_data.csv") 
EDA <- read.csv(text = EDA_GIT_LINK)

# going to remove columns 11,12 since they are geographical data so pretty useless right now

EDA <- EDA[,-c(11,12)]

# make all var undercase

EDA <- clean_names(EDA)

# renaming response variables to better names

colnames(EDA)[7] <- "presence_absence_sch"
colnames(EDA)[8] <- "density_sch"
colnames(EDA)[9] <- "presence_absence_typ"
colnames(EDA)[10] <- "density_typ"
names(EDA)
```

# Structure Analysis

```{r}
#| echo: false

# Changing columns 1,2, 7-15

EDA[, c(1,2,7:15)] <- lapply(EDA[, c(1,2,7:15)], function(x) gsub("[^0-9.]", "", x))
EDA[, c(1,2,7:15)] <- lapply(EDA[, c(1,2,7:15)], as.numeric)

str(EDA)

unique(EDA$transect) 
unique(EDA$site_combined) 
unique(EDA$site_individual)
unique(EDA$exposure)

summary(EDA$presence_absence_sch)
summary(EDA$presence_absence_typ)
summary(EDA$density_sch)
summary(EDA$density_typ)
```

-   There are 24 unique transects

-   4 unique sites, 8 individual sites

## NA Assess

```{r}
#| echo: false 
rows_with_na <- EDA[!complete.cases(EDA), ] 
print(rows_with_na)  
unique(rows_with_na$site_combined)
EDA <- na.omit(EDA)
```

-   Only one missing variable (TYP_D) in obs. 289

## Summarizing Response Variables

```{r}
#| echo: false
names(EDA)
EDA$presence_absence <- (EDA$presence_absence_sch + EDA$presence_absence_typ)
EDA$density <- (EDA$density_sch + EDA$density_typ)

EDA <- EDA %>%
  mutate(plant_type = case_when(
    presence_absence_typ == 1 ~ "typ",
    presence_absence_sch == 1 ~ "sch",
    TRUE ~ NA_character_
  ))

summary(EDA$presence_absence_sch)
summary(EDA$presence_absence_typ)
summary(EDA$presence_absence)

summary(EDA$density_sch)
summary(EDA$density_typ)
summary(EDA$density)
```

# Distribution of Explanatory Variables

## Cost (predicted intensity of wave)

```{r}
ggplot(EDA, aes(x = cost_small, fill = exposure)) +
  geom_histogram(binwidth = 1, color = "black", alpha = 0.7, position = "dodge") +
  labs(title = "Distribution of Cost (predicted intensity of waves)", 
       x = "Cost", y = "Frequency") +
  theme_minimal()
```

## Fetch Projected

```{r}
#| echo: false
ggplot(EDA, aes(x = fetch_projected, fill = exposure)) +
  geom_histogram(binwidth = 100, color = "black", alpha = 0.7, position = "Identity") +
  labs(title = "Distribution of Fetch Projected", x = "Fetch Projected", y = "Frequency") +
  theme_minimal()
```

## Water Depth

```{r}
#| echo: false
ggplot(EDA, aes(x = water_depth, fill = exposure)) +
  geom_histogram(binwidth = 5, color = "black", alpha = 0.7, position = "Dodge") +
  labs(title = "Distribution of Water Depth", x = "Water Depth", y = "Frequency") +
  theme_minimal()
```

-   Fairly normal-ish

## Slope

```{r}
#| echo: false 
EDA$slope_use <- abs(EDA$slope_use)

ggplot(EDA, aes(x = slope_use, fill = exposure)) +
  geom_histogram(binwidth = 1, color = "black", alpha = 0.7, position = "Identity") +
  labs(title = "Distribution of Slope", x = "Slope", y = "Frequency") +
  theme_minimal()
```

-   Has a similar distribution to the cost variable

## Speed Limit

```{r}
#| echo: false 
ggplot(EDA, aes(x = speed_limit)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Speed Limit", x = "Speed Limit", y = "Frequency") +
  theme_minimal()
```

# Exposure type

## Cost

### Presence/Absence Against Cost

```{r}
#| echo: false

ggplot(EDA, aes(x = cost_small, y = presence_absence, col = exposure)) +
  geom_point(size = 3, alpha = 0.7) +
  facet_wrap(vars(exposure)) +
  labs(title = "Presence/Absence (SCH) Against Cost Grouped By Exposure", 
       x = "Cost", 
       y = "Presence/Absence") +
  theme_minimal()
```

### Density Against Cost

```{r}
#| echo: false 

ggplot(EDA, aes(x = cost_small, y = density, col = exposure)) +
  geom_point(size = 3, alpha = 0.7) +
  facet_wrap(vars(exposure)) +
  labs(title = "Density Against Cost Grouped By Exposure",
       x = "Cost", 
       y = "Density") +
  theme_minimal()
```

## Fetch Project

### Presence/Absence Against Fetch

```{r}
#| echo: false 

ggplot(EDA, aes(x = fetch_projected, y = presence_absence, col = exposure)) + facet_wrap(vars(exposure)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Presence/Absence Against Fetch Grouped By Exposure", 
       x = "Fetch", 
       y = "Presence/Absence") +
  theme_minimal()
```

### Density Against Fetch

```{r}
#| echo: false 

ggplot(EDA, aes(x = fetch_projected, y = density, col = exposure)) +
  facet_wrap(vars(exposure)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Density Against Fetch Grouped By Exposure", 
       x = "Fetch", 
       y = "Density") +
  theme_minimal()
```

## Water Depth

### Presence/Absence Against Water Depth

```{r}
#| echo: false

ggplot(EDA, aes(x = water_depth, y = presence_absence, col = exposure)) +
  facet_wrap(vars(exposure)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Presence/Absence Against Water Depth Grouped By Exposure", 
       x = "Water Depth", 
       y = "Presence/Absence") +
  theme_minimal()
```

### Density Against Water Depth

```{r}
#| echo: false 
ggplot(EDA, aes(x = water_depth, y = density, col = exposure)) +
  facet_wrap(vars(exposure)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Density Against Water Depth Grouped By Exposure", 
       x = "Water Depth", 
       y = "Density") +
  theme_minimal()
```

## Slope

### Presence/Absence Against Slope

```{r}
#| echo: false 

ggplot(EDA, aes(x = slope_use, y = presence_absence, col = exposure)) +
  facet_wrap(vars(exposure)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Presence/Absence Against Slope Grouped By Exposure", 
       x = "Slope", 
       y = "Presence/Absence") +
  theme_minimal()

```

### Density Against Slope

```{r}
#| echo: false 

ggplot(EDA, aes(x = slope_use, y = density, col = exposure)) +
  facet_wrap(vars(exposure)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Density Against Slope Grouped By Exposure", 
       x = "Slope", 
       y = "Density") +
  theme_minimal()
```

## Speed Limit

### Presence/Absence Against Speed Limit (Bar Plot)

```{r}
#| echo: false 

ggplot(EDA, aes(x = factor(speed_limit), fill = factor(presence_absence), col = exposure)) +
  facet_wrap(vars(exposure)) +
  geom_bar(position = "dodge") +
  labs(title = "Presence/Absence by Speed Limit Grouped By Exposure",
       x = "Speed Limit",
       y = "Frequency",
       fill = "Presence/Absence") +
  theme_minimal()
```

### Density Against Speed Limit

```{r}
#| echo: false 

ggplot(EDA, aes(x = speed_limit, y = density, col = exposure)) +
  facet_wrap(vars(exposure)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Density Against Speed Limit Grouped By Exposure", 
       x = "Speed Limit", 
       y = "Density") +
  theme_minimal()

```

# Site Type

## Cost

### Presence/Absence Against Cost

```{r}
#| echo: false

ggplot(EDA, aes(x = cost_small, y = presence_absence, col = site_combined)) +
  geom_point(size = 3, alpha = 0.7) +
  facet_wrap(vars(site_combined)) +
  labs(title = "Presence/Absence Against Cost Grouped By Site", 
       x = "Cost", 
       y = "Presence/Absence") +
  theme_minimal()
```

### Density Against Cost

```{r}
#| echo: false 

ggplot(EDA, aes(x = cost_small, y = density, col = site_combined)) +
  geom_point(size = 3, alpha = 0.7) +
  facet_wrap(vars(site_combined)) +
  labs(title = "Density Against Cost Grouped By Site",
       x = "Cost", 
       y = "Density") +
  theme_minimal()

ggplot(EDA, aes(x = cost_small, y = density, col = site_combined)) +
  geom_point(size = 3, alpha = 0.7) +
  facet_wrap(vars(exposure)) +
  labs(title = "Density Against Cost Grouped By Site",
       x = "Cost", 
       y = "Density") +
  theme_minimal()
```

## Fetch Project

### Presence/Absence Against Fetch

```{r}
#| echo: false 

ggplot(EDA, aes(x = fetch_projected, y = presence_absence, col = site_combined)) + facet_wrap(vars(site_combined)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Presence/Absence Against Fetch Grouped By Site", 
       x = "Fetch", 
       y = "Presence/Absence") +
  theme_minimal()
```

### Density Against Fetch

```{r}
#| echo: false 

ggplot(EDA, aes(x = fetch_projected, y = density, col = site_combined)) +
  facet_wrap(vars(site_combined)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Density Against Fetch Grouped By Site", 
       x = "Fetch", 
       y = "Density") +
  theme_minimal()
```

## Water Depth

### Presence/Absence Against Water Depth

```{r}
#| echo: false

ggplot(EDA, aes(x = water_depth, y = presence_absence, col = site_combined)) +
  facet_wrap(vars(site_combined)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Presence/Absence Against Water Depth Grouped By Site", 
       x = "Water Depth", 
       y = "Presence/Absence") +
  theme_minimal()
```

### Density Against Water Depth

```{r}
#| echo: false 

ggplot(EDA, aes(x = water_depth, y = density, col = site_combined)) +
  facet_wrap(vars(site_combined)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Density Against Water Depth Grouped By Site", 
       x = "Water Depth", 
       y = "Density") +
  theme_minimal()
```

## Slope

### Presence/Absence Against Slope

```{r}
#| echo: false 

ggplot(EDA, aes(x = slope_use, y = presence_absence, col = site_combined)) +
  facet_wrap(vars(site_combined)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Presence/Absence Against Slope Grouped By Site", 
       x = "Slope", 
       y = "Presence/Absence") +
  theme_minimal()
```

### Density Against Slope

```{r}
#| echo: false 

ggplot(EDA, aes(x = slope_use, y = density, col = site_combined)) +
  facet_wrap(vars(site_combined)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Density Against Slope Grouped By Site", 
       x = "Slope", 
       y = "Density") +
  theme_minimal()
```

## Speed Limit

### Presence/Absence Against Speed Limit (Bar Plot)

```{r}
#| echo: false 

ggplot(EDA, aes(x = factor(speed_limit), fill = factor(presence_absence), col = site_combined)) +
  facet_wrap(vars(site_combined)) +
  geom_bar(position = "dodge") +
  labs(title = "Presence/Absence by Speed Limit Grouped By Site",
       x = "Speed Limit",
       y = "Frequency",
       fill = "Presence/Absence") +
  theme_minimal()
```

### Density Against Speed Limit

```{r}
#| echo: false 

ggplot(EDA, aes(x = speed_limit, y = density, col = site_combined)) +
  facet_wrap(vars(site_combined)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Density Against Speed Limit Grouped By Site", 
       x = "Speed Limit", 
       y = "Density") +
  theme_minimal()
```

# Plant type

## Cost

### Density Against Cost

```{r}
#| echo: false 

EDA_filtered2 <- na.omit(EDA)

ggplot(EDA_filtered2, aes(x = cost_small, y = density, col = plant_type)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Density Against Cost Grouped By Plant Type",
       x = "Cost", 
       y = "Density") +
  theme_minimal()
```

## Fetch Project

### Density Against Fetch

```{r}
#| echo: false 

ggplot(EDA_filtered2, aes(x = fetch_projected, y = density, col = plant_type)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Density Against Fetch Grouped By Plant Type", 
       x = "Fetch", 
       y = "Density") +
  theme_minimal()
```

## Water Depth

### Density Against Water Depth

```{r}
#| echo: false 

ggplot(EDA_filtered2, aes(x = water_depth, y = density, col = plant_type)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Density Against Water Depth Grouped By Plant Type", 
       x = "Water Depth", 
       y = "Density") +
  theme_minimal()
```

## Slope

### Density Against Slope

```{r}
#| echo: false 

ggplot(EDA_filtered2, aes(x = slope_use, y = density, col = plant_type)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Density Against Slope Grouped By Plant Type", 
       x = "Slope", 
       y = "Density") +
  theme_minimal()
```
