---
title: "STAT-370-PROJ-EDA" 
author: "William Quinn" 
format: html
self-contained: true
editor: visual
---

# Libraries {.appendix}

```{r}
library(RCurl)
library(ggplot2)
library(janitor)
library(dplyr)
```

# Importing {.appendix}

```{r}
EDA_GIT_LINK <- getURL("https://raw.githubusercontent.com/Cgrossman23/STAT-370-PROJ/refs/heads/main/cleaned_data.csv") 
EDA <- read.csv(text = EDA_GIT_LINK)

# going to remove columns 11,12 since they are geographical data so pretty useless right now

EDA <- EDA[,-c(11,12)]

# make all var undercase

EDA <- clean_names(EDA)
```

# EDA

## Structure Analysis

```{r}
#| echo: false 
str(EDA) 
unique(EDA$transect) 
unique(EDA$site_combined) 
unique(EDA$site_individual) 
summary(EDA$speed_limit)
```

-   There are 24 unique transects. What are these?

-   4 unique sites, 8 individual sites

-   Speed limit has little variable, 10 or 14 mph

## NA Assess

```{r}
#| echo: false 
rows_with_na <- EDA[!complete.cases(EDA), ] 
print(rows_with_na)  
unique(rows_with_na$site_combined)
```

-   Only one missing variable (TYP_D) in obs. 289

## Distribution Visualizations of Explanatory Variables

### Cost (predicted intensity of waves)

```{r}
ggplot(EDA, aes(x = cost_small)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Cost (predicted intensity of waves)", x = "Cost", y = "Frequency") +
  theme_minimal()
```

### Water Depth

```{r}
#| echo: false
ggplot(EDA, aes(x = water_depth)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Water Depth", x = "Water Depth", y = "Frequency") +
  theme_minimal()
```

-   Fairly normal-ish

### Fetch Projected

```{r}
#| echo: false
ggplot(EDA, aes(x = fetch_projected)) +
  geom_histogram(binwidth = 100, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Fetch Projected", x = "Fetch Projected", y = "Frequency") +
  theme_minimal()
```

-   Fairly uniform

### Slope

```{r}
#| echo: false 
EDA$slope_use <- abs(EDA$slope_use)

ggplot(EDA, aes(x = slope_use)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Slope", x = "Slope", y = "Frequency") +
  theme_minimal()
```

-   Has a similar distribution to the cost variable

### Speed Limit

```{r}
#| echo: false 
ggplot(EDA, aes(x = speed_limit)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Speed Limit", x = "Speed Limit", y = "Frequency") +
  theme_minimal()
```

## Scatter Plots Response Variable Against Explanatory

```{r}

```
