---
title: "STAT-370-PROJ-MODELING"
author: "William Quinn"
format: html
editor: visual
---

# Libraries {.appendix}

```{r}
library(RCurl)
library(janitor)
library(dplyr)
library(car)
library(flexplot)
library(MASS)
library(ggplot2)

#mixed model libraries
library(lme4)
library(HLMdiag)
library(flexplot)
library(MuMIn)
library(glmmTMB) 

#geographic regression libraries
library(spdep)
library(deldir)

#for zero inflated
library(pscl)
```

# Importing {.appendix}

```{r}
data_link <- getURL("https://raw.githubusercontent.com/Cgrossman23/STAT-370-PROJ/refs/heads/main/cleaned_data.csv") 
wave <- read.csv(text = EDA_GIT_LINK)
wave <- as.data.frame(wave)


# make all var undercase

wave <- clean_names(wave)

# renaming response variables to better names

colnames(wave)[7] <- "presence_absence_sch"
colnames(wave)[8] <- "density_sch"
colnames(wave)[9] <- "presence_absence_typ"
colnames(wave)[10] <- "density_typ"
names(wave)
```

# Structure Reference

```{r}
#| echo: false
names(wave)

wave[, c(1,2,7:15)] <- lapply(wave[, c(1,2,7:15)], function(x) gsub("[^0-9.]", "", x))
wave[, c(1,2,7:15)] <- lapply(wave[, c(1,2,7:15)], as.numeric)

str(wave)

unique(wave$transect) 
unique(wave$site_combined) 
unique(wave$site_individual)
unique(wave$exposure)
```

## NA Assess

```{r}
#| echo: false 
rows_with_na <- wave[!complete.cases(wave), ] 
print(rows_with_na)  
unique(rows_with_na$site_combined)
wave <- na.omit(wave)
```

## Summarizing Response Variables

```{r}
#| echo: false 

wave$presence_absence <- (wave$presence_absence_sch + wave$presence_absence_typ)
wave$density <- (wave$density_sch + wave$density_typ)

wave <- wave %>% mutate(plant_type = case_when( 
                presence_absence_typ ==  1 ~ "typ", 
                presence_absence_sch == 1 ~ "sch", 
                TRUE ~ NA_character_ ))
```

## Filter for Unexposed and Site

```{r}
wave$longitude <- ifelse(wave$longitude > 0, -wave$longitude, wave$longitude)

wave2 <- wave %>% filter(exposure == "Unexposed")
wave3 <- wave %>% filter(exposure == "Exposed")

wave_a <- wave2 %>% filter(site_combined == "A")
wave_b <- wave2 %>% filter(site_combined == "B")
wave_c <- wave2 %>% filter(site_combined == "C")
wave_d <- wave2 %>% filter(site_combined == "D")

wave_a2 <- wave3 %>% filter(site_combined == "A")
wave_b2 <- wave3 %>% filter(site_combined == "B")
wave_c2 <- wave3 %>% filter(site_combined == "C")
wave_d2 <- wave3 %>% filter(site_combined == "D")
```

# Geographical Correlation

## Site A - unexposed

```{r}
coords_a <- cbind(wave_a$latitude, wave_a$longitude)

distances <- seq(0.000001, 0.001, by = 0.000001)
values <- numeric(length(distances))
values[] <- NA  
mmi_a <- -Inf  
od_a <- NA
neighbors_a <- NULL
weights_a <- NULL
options(warn = -1)
for (i in seq_along(distances)) {
  neighbors <- dnearneigh(coords_a, d1 = 0, d2 = distances[i]) #insert cords
  if (sum(card(neighbors)) == 0) { next }
  dists <- nbdists(neighbors, coords_a) # and here
  weights <- nb2listw(neighbors, 
                        glist = lapply(dists, function(x) 1 / x), 
                        style = "W", 
                        zero.policy = TRUE)
  moran_results <- moran(wave_a$density, weights, length(neighbors), Szero(weights)) #insert density here
  values[i] <- moran_results$I
  
  if (!is.na(values[i]) && values[i] > mmi_a) {
    mmi_a <- values[i]
    od_a <- distances[i]
    neighbors_a <- neighbors
    weights_a <- weights
  }
}
cat("Maximum Moran's I:", mmi_a, "\n")
cat("Optimal Distance:", od_a, "\n")

plot(neighbors_a,coords_a, col = "red")
```

## Site B - unexposed

```{r}
coords_b <- cbind(wave_b$latitude, wave_b$longitude)
  
distances <- seq(0.00012, 0.001, by = 0.000001)
values <- numeric(length(distances))
values[] <- NA  
mmi_b <- -Inf  
od_b <- NA
neighbors_b <- NULL
weights_b <- NULL
options(warn = -1)
for (i in seq_along(distances)) {
  neighbors <- dnearneigh(coords_b, d1 = 0, d2 = distances[i]) #insert cords
  if (sum(card(neighbors)) == 0) { next }
  dists <- nbdists(neighbors, coords_b) # and here
  weights <- nb2listw(neighbors, 
                        glist = lapply(dists, function(x) 1 / x), 
                        style = "W", 
                        zero.policy = TRUE)
  moran_results <- moran(wave_b$density, weights, length(neighbors), Szero(weights)) #insert density here
  values[i] <- moran_results$I
  
  if (!is.na(values[i]) && values[i] > mmi_b) {
    mmi_b <- values[i]
    od_b <- distances[i]
    neighbors_b <- neighbors
    weights_b <- weights
  }
}
cat("Maximum Moran's I:", mmi_b, "\n")
cat("Optimal Distance:", od_b, "\n")

plot(neighbors_b,coords_b, col = "red")
```

## Site C - unexposed

```{r}
coords_c <- cbind(wave_c$latitude, wave_c$longitude)
  
distances <- seq(0.000001, 0.001, by = 0.000001)
values <- numeric(length(distances))
values[] <- NA  
mmi_c <- -Inf  
od_c <- NA
neighbors_c <- NULL
weights_c <- NULL
options(warn = -1)
for (i in seq_along(distances)) {
  neighbors <- dnearneigh(coords_c, d1 = 0, d2 = distances[i]) #insert cords
  if (sum(card(neighbors)) == 0) { next }
  dists <- nbdists(neighbors, coords_c) # and here
  weights <- nb2listw(neighbors, 
                        glist = lapply(dists, function(x) 1 / x), 
                        style = "W", 
                        zero.policy = TRUE)
  moran_results <- moran(wave_c$density, weights, length(neighbors), Szero(weights)) #insert density here
  values[i] <- moran_results$I
  
  if (!is.na(values[i]) && values[i] > mmi_c) {
    mmi_c <- values[i]
    od_c <- distances[i]
    neighbors_c <- neighbors
    weights_c <- weights
  }
}
cat("Maximum Moran's I:", mmi_c, "\n")
cat("Optimal Distance:", od_c, "\n")

plot(neighbors_c,coords_c, col = "red")
```

## Site D - unexposed

```{r}
coords_d <- cbind(wave_d$latitude, wave_d$longitude)

distances <- seq(0.000001, 0.001, by = 0.000001)
values <- numeric(length(distances))
values[] <- NA  
mmi_d <- -Inf  
od_d <- NA
neighbors_d <- NULL
weights_d <- NULL
options(warn = -1)
for (i in seq_along(distances)) {
  neighbors <- dnearneigh(coords_d, d1 = 0, d2 = distances[i]) #insert cords
  if (sum(card(neighbors)) == 0) { next }
  dists <- nbdists(neighbors, coords_d) # and here
  weights <- nb2listw(neighbors, 
                        glist = lapply(dists, function(x) 1 / x), 
                        style = "W", 
                        zero.policy = TRUE)
  moran_results <- moran(wave_d$density, weights, length(neighbors), Szero(weights)) #insert density here
  values[i] <- moran_results$I
  
  if (!is.na(values[i]) && values[i] > mmi_d) {
    mmi_d <- values[i]
    od_d <- distances[i]
    neighbors_d <- neighbors
    weights_d <- weights
  }
}
cat("Maximum Moran's I:", mmi_d, "\n")
cat("Optimal Distance:", od_d, "\n")

plot(neighbors_d,coords_d, col = "red")
```

## Site A - exposed

```{r}
coords_a2 <- cbind(wave_a2$latitude, wave_a2$longitude)

n_a2 <- knn2nb(knearneigh(coords_a2, k=4, longlat = TRUE))
wa <- nb2listw(n_a2, style = "W", zero.policy = TRUE)
mra <- moran(wave_a2$density, wa, length(n_a2), Szero(wa)) 

print(wa)
head(mra)

plot(n_a2, coords_a2, col = "red")
```

## Site B - exposed

```{r}
coords_b2 <- cbind(wave_b2$latitude, wave_b2$longitude)

n_b2 <- knn2nb(knearneigh(coords_b2, k=3, longlat = TRUE))
wb <- nb2listw(n_b2, style = "W", zero.policy = TRUE)
mrb <- moran(wave_b2$density, wb, length(n_b2), Szero(wb)) 

print(wb)
head(mrb)

plot(n_b2, coords_b2, col = "red")
```

## Site C - exposed

```{r}
coords_c2 <- cbind(wave_c2$latitude, wave_c2$longitude)


n_c2 <- knn2nb(knearneigh(coords_c2, k=3, longlat = TRUE))
wc <- nb2listw(n_c2, style = "W", zero.policy = TRUE)
mrc <- moran(wave_b2$density, wc, length(n_c2), Szero(wc)) 

print(wc)
head(mrc)

plot(n_c2, coords_c2, col = "red")
```

## Site D - exposed

```{r}
coords_d2 <- cbind(wave_c2$latitude, wave_c2$longitude)


n_d2 <- knn2nb(knearneigh(coords_d2, k=3, longlat = TRUE))
wd <- nb2listw(n_d2, style = "W", zero.policy = TRUE)
mrd <- moran(wave_c2$density, wd, length(n_d2), Szero(wd)) 

print(wd)
head(mrd)

plot(n_d2, coords_d2, col = "red")
```

# Geographic correlation vs ZIM

A **zero-inflated model** assumes the observed count data is generated by a mixture of:

-   A **binary process** that determines whether an observation is in the "always-zero" group.

-   A **count process** (e.g., Poisson or Negative Binomial) that generates nonzero counts and some additional zeros.

    $$
    P(Y=0)=\pi+(1-\pi)e^{-\lambda}
    $$

$$
P(Y=k)=\pi+(1-\pi)\dfrac{\lambda^{k}e^{-\lambda}}{k!},\ k\gt 0
$$

where:

-   $\pi$ is the probability of a **structural zero**.

-   $\lambda$ is the **Poisson mean** for the count-generating process.

## ZIM

### Site A

```{r}
table(wave_a$density)

unique(sort(round(log(wave_a$density+1))))
unique(sort(round(log(wave_a$density+1)*10)))
unique(sort(wave_a$density))

wave_a$log_density <- round(log(wave_a$density+1)*10)

model_zero_a <- zeroinfl(log_density ~ fetch_projected , data = wave_a, dist="negbin")

estimates(model_zero_a)

resid_a <- residuals(model_zero_a)

moran_resid_test <- moran.test(resid_a, weights_a)

print(moran_resid_test) # significant
```

### Site B

```{r}
table(wave_b$density)

unique(sort(round(log(wave_b$density+1))))
unique(sort(round(log(wave_b$density+1)*10)))
unique(sort(wave_b$density))

wave_b$log_density <- round(log(wave_b$density+1)*10)

model_zero_b <- zeroinfl(log_density ~ fetch_projected , data = wave_b, dist="negbin")

estimates(model_zero_b)

resid_b <- residuals(model_zero_b)

moran_resid_test <- moran.test(resid_b, weights_b)

print(moran_resid_test) # insignificant
```

### Site C

```{r}
table(wave_c$density)

unique(sort(round(log(wave_c$density+1))))
unique(sort(round(log(wave_c$density+1)*10)))
unique(sort(wave_c$density))

wave_c$log_density <- round(log(wave_c$density+1)*10)

model_zero_c <- zeroinfl(log_density ~ fetch_projected , data = wave_c, dist="negbin")

estimates(model_zero_c)

resid_c <- residuals(model_zero_c)

moran_resid_test <- moran.test(resid_c, weights_c)

print(moran_resid_test) # significant
```

### Site D

```{r}
table(wave_d$density)

unique(sort(round(log(wave_d$density+1))))
unique(sort(round(log(wave_d$density+1)*10)))
unique(sort(wave_d$density))

wave_d$log_density <- round(log(wave_d$density+1)*10)

m0d <- zeroinfl(log_density ~ fetch_projected , data = wave_d, dist="negbin")
m0d1 <- zeroinfl(log_density ~ fetch_projected + water_depth , data = wave_d, dist="negbin")
m0d2 <- zeroinfl(log_density ~ fetch_projected + water_depth + slope_use , data = wave_d, dist="negbin")

AIC(m0d)
AIC(m0d1)
AIC(m0d2)

logLik(m0d)
logLik(m0d1)
logLik(m0d2)

vuong(m0d, m0d1)

estimates(m0d1)

resid_d <- residuals(m0d1)

moran_resid_test <- moran.test(resid_d, weights_d)

print(moran_resid_test) # insignificant
```

## Mixed Effect Model

### Baseline - site_combined

```{r}
model1 <- lmer(density ~
                        1
                        + (1 | site_combined),
                        data = wave)

r_squared <- r.squaredGLMM(model1)
print(r_squared)
# 1.48% of variance due to clustering between sites
```

### Baseline - exposure nested in site_combined

```{r}
model2 <- lmer(density ~
                        1
                        + (1 | exposure)
                        + (1 | site_combined),
                        data = wave)

r_squared2 <- r.squaredGLMM(model2)
print(r_squared2)
# 21.66% of variance explained

anova(model2, model1)
# nested model is better
```

### Exp - fetch_projected with exposure nested in site_combined

```{r}
model3 <- lmer(density ~
                        fetch_projected
                        + (1 | exposure)
                        + (1 | site_combined),
                        data = wave)

r_squared3 <- r.squaredGLMM(model3)
print(r_squared3)
# basically zero explanation from fetch

anova(model3, model2)
# baseline is better
```

### Exp - water_depth and fetch_projected with exposure nested in site_combined

```{r}
model4 <- lmer(density ~
                        fetch_projected
                        + water_depth
                        + (1 | exposure)
                        + (1 | site_combined),
                        data = wave)

r_squared4 <- r.squaredGLMM(model4)
print(r_squared4)
# 2.09% of variance explained by water_depth

anova(model4, model2)
# close
```

### Exp - water_depth\*fetch_projected with exposure nested in site_combined

```{r}
model5 <- lmer(density ~
                        fetch_projected*water_depth
                        + (1 | exposure)
                        + (1 | site_combined),
                        data = wave)

r_squared5 <- r.squaredGLMM(model5)
print(r_squared5)
# 10.03% of variance explained by water_depth*fetch_projected

anova(model5, model2)
anova(model5, model4)
# interaction model is better
```

### ZIM - nested no interaction

```{r}
model01 <- glmmTMB(density ~ fetch_projected + water_depth 
                            + (1 | exposure)
                            + (1 | site_combined),
                    ziformula = ~ 1, 
                    family = "nbinom2",
                    data = wave)
summary(model01)
r.squaredGLMM(model01)
# higher explanatory power in all cases than regular MEM

AIC(model5)
AIC(model01)
# much better
```

### ZIM - nested interaction

```{r}
model02 <- glmmTMB(density ~ fetch_projected*water_depth 
                            + (1 | exposure)
                            + (1 | site_combined),
                    ziformula = ~ 1, 
                    family = "nbinom2",
                    data = wave)
summary(model02)
r.squaredGLMM(model02)

anova(model02, model01)
# worse
```

### ZIM - nested just water_depth

```{r}
model03 <- glmmTMB(density ~ water_depth 
                            + (1 | exposure)
                            + (1 | site_combined),
                    ziformula = ~ 1, 
                    family = "nbinom2",
                    data = wave)
summary(model03)

r.squaredGLMM(model03)
r.squaredGLMM(model01)
# model 3 is better

anova(model03, model01)
#close but model 3 is better?
```

# LISA Model

```{r}
coords <- cbind(wave$latitude, wave$longitude)
head(wave$coords)
distances <- seq(0.00012, 0.001, by = 0.000001)
values <- numeric(length(distances))
values[] <- NA  
mmi <- -Inf  
od <- NA
neighbors <- NULL
weights <- NULL
options(warn = -1)
for (i in seq_along(distances)) {
  neighbors <- dnearneigh(coords, d1 = 0, d2 = distances[i]) #insert cords
  if (sum(card(neighbors)) == 0) { next }
  dists <- nbdists(neighbors, coords) # and here
  weights <- nb2listw(neighbors, 
                        glist = lapply(dists, function(x) 1 / x), 
                        style = "W", 
                        zero.policy = TRUE)
  moran_results <- moran(wave$density, weights, length(neighbors), Szero(weights)) #insert density here
  values[i] <- moran_results$I
  
  if (!is.na(values[i]) && values[i] > mmi) {
    mmi <- values[i]
    od <- distances[i]
    neighbors <- neighbors
    weights <- weights
  }
}
cat("Maximum Moran's I:", mmi, "\n")
cat("Optimal Distance:", od, "\n")

plot(neighbors,coords, col = "red")

# Compute Moran’s I
moran_test <- moran.test(wave$density, listw = weights, zero.policy = TRUE)
print(moran_test)

# Global Moran's I is highly significant.

# Comparison to best model found before
resid <- residuals(model03)

moran_resid_test <- moran.test(resid, weights)

print(moran_resid_test) # significant

# But we know that local clustering effects are likely important as we proved for each site
```

```{r}
lmr_a <- localmoran(wave_a$density, weights_a, zero.policy = TRUE)
lmr_a <- lmr_a[, c(1,5)]
lmr_a[is.na(lmr_a)] <- 0
wave_a2 <-cbind(wave_a, lmr_a)
print(wave_a2)

lmr_b <- localmoran(wave_b$density, weights_b, zero.policy = TRUE)
lmr_b <- lmr_b[, c(1,5)]
lmr_b[is.na(lmr_b)] <- 0
wave_b2 <-cbind(wave_b, lmr_b)
print(wave_b2)

lmr_c <- localmoran(wave_c$density, weights_c, zero.policy = TRUE)
lmr_c <- lmr_c[, c(1,5)]
lmr_c[is.na(lmr_c)] <- 0
wave_c2 <-cbind(wave_c, lmr_c)
print(wave_c2)

lmr_d <- localmoran(wave_d$density, weights_d, zero.policy = TRUE)
lmr_d <- lmr_d[, c(1,5)]
lmr_d[is.na(lmr_d)] <- 0
wave_d2 <- cbind(wave_d, lmr_d)
print(wave_d2)

wave3 <-rbind(wave_a2, wave_b2, wave_c2, wave_d2)
```

```{r}
library(Matrix)

weights_list <- list(weights_a, weights_b, weights_c, weights_d, wa, wb, wc, wd)

sparse_weights <- lapply(weights_list, function(w) {
  mat <- listw2mat(w)
  as(mat, "CsparseMatrix")
})

combined_sparse <- bdiag(sparse_weights)

combined_neighbors <- mat2listw(combined_sparse, style = "W", zero.policy = TRUE)

resid <- residuals(model03)

length(resid)
length(combined_neighbors)

moran_result <- moran.test(resid, combined_neighbors)

print(moran_result)
```

```{r}
library(spatialreg)

# Create a new spatial weights matrix for the merged dataset
merged_coords <- cbind(final_dataset$latitude, final_dataset$longitude)
merged_neighbors <- dnearneigh(merged_coords, d1 = 0, d2 = max(final_dataset$optimal_distance), zero.policy = TRUE)
merged_weights <- nb2listw(merged_neighbors, style = "W", zero.policy = TRUE)

# Run a Spatial Lag Model
formula <- density ~ independent_var1 + independent_var2
slm_model <- lagsarlm(formula, data = final_dataset, listw = merged_weights, zero.policy = TRUE)

summary(slm_model)

```
