---
title: "STAT-370-PROJ-EDA" 
author: "William Quinn" 
format: html
self-contained: true
editor: visual
---

# Libraries {.appendix}

```{r}
library(RCurl)
library(ggplot2)
library(janitor)
library(dplyr)
```

# Importing {.appendix}

```{r}
EDA_GIT_LINK <- getURL("https://raw.githubusercontent.com/Cgrossman23/STAT-370-PROJ/refs/heads/main/cleaned_data.csv") 
EDA <- read.csv(text = EDA_GIT_LINK)

# going to remove columns 11,12 since they are geographical data so pretty useless right now

EDA <- EDA[,-c(11,12)]

# make all var undercase

EDA <- clean_names(EDA)

# renaming response variables to better names

colnames(EDA)[7] <- "presence_absence_sch"
colnames(EDA)[8] <- "density_sch"
colnames(EDA)[9] <- "presence_absence_typ"
colnames(EDA)[10] <- "density_typ"
names(EDA)

```

# EDA

##  Structure Analysis

```{r}
#| echo: false 
# Changing columns 1,2, 7-15
EDA[, c(1,2,7:15)] <- lapply(EDA[, c(1,2,7:15)], function(x) gsub("[^0-9.]", "", x))
EDA[, c(1,2,7:15)] <- lapply(EDA[, c(1,2,7:15)], as.numeric)

str(EDA)

unique(EDA$transect) 
unique(EDA$site_combined) 
unique(EDA$site_individual)

summary(EDA$presence_absence_sch)
summary(EDA$presence_absence_typ)
summary(EDA$density_sch)
summary(EDA$density_typ)
```

-   There are 24 unique transects

-   4 unique sites, 8 individual sites

## NA Assess

```{r}
#| echo: false 
rows_with_na <- EDA[!complete.cases(EDA), ] 
print(rows_with_na)  
unique(rows_with_na$site_combined)
```

-   Only one missing variable (TYP_D) in obs. 289

## Distribution Visualizations of Explanatory Variables

### Cost (predicted intensity of waves)

```{r}
ggplot(EDA, aes(x = cost_small)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Cost (predicted intensity of waves)", x = "Cost", y = "Frequency") +
  theme_minimal()
```

### Fetch Project

```{r}
#| echo: false
ggplot(EDA, aes(x = fetch_projected)) +
  geom_histogram(binwidth = 100, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Fetch Projected", x = "Fetch Projected", y = "Frequency") +
  theme_minimal()
```

### Water Depth

```{r}
#| echo: false
ggplot(EDA, aes(x = water_depth)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Water Depth", x = "Water Depth", y = "Frequency") +
  theme_minimal()
```

-   Fairly normal-ish

### Slope

```{r}
#| echo: false 
EDA$slope_use <- abs(EDA$slope_use)

ggplot(EDA, aes(x = slope_use)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Slope", x = "Slope", y = "Frequency") +
  theme_minimal()
```

-   Has a similar distribution to the cost variable

### Speed Limit

```{r}
#| echo: false 
ggplot(EDA, aes(x = speed_limit)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Speed Limit", x = "Speed Limit", y = "Frequency") +
  theme_minimal()
```

## Scatter Plots Response Variable Against Explanatory

### Cost

#### Presence/Absence Against Cost

```{r}
#| echo: false 

ggplot(EDA, aes(x = cost_small, y = presence_absence_sch)) +
  geom_point(color = "red", size = 3, alpha = 0.7) +
  labs(title = "Scatterplot of Presence/Absence (SCH) Against Cost", 
       x = "Cost", 
       y = "Presence/Absence") +
  theme_minimal()

ggplot(EDA, aes(x = cost_small, y = presence_absence_typ)) +
  geom_point(color = "red", size = 3, alpha = 0.7) +
  labs(title = "Scatterplot of Presence/Absence (TYP) Against Cost", 
       x = "Cost", 
       y = "Presence/Absence") +
  theme_minimal()
```

#### Density Against Cost

```{r}
#| echo: false 

ggplot(EDA, aes(x = cost_small, y = density_sch)) +
  geom_point(color = "red", size = 3, alpha = 0.7) +
  labs(title = "Scatterplot of Density (SCH) Against Cost", 
       x = "Cost", 
       y = "Density") +
  theme_minimal()

ggplot(EDA, aes(x = cost_small, y = density_typ)) +
  geom_point(color = "red", size = 3, alpha = 0.7) +
  labs(title = "Scatterplot of Density (TYP) Against Cost", 
       x = "Cost", 
       y = "Density") +
  theme_minimal()
```

### Fetch Project

#### Presence/Absence Against Fetch

```{r}
#| echo: false 

ggplot(EDA, aes(x = fetch_projected, y = presence_absence_sch)) +
  geom_point(color = "red", size = 3, alpha = 0.7) +
  labs(title = "Scatterplot of Presence/Absence (SCH) Against Fetch", 
       x = "Fetch", 
       y = "Presence/Absence") +
  theme_minimal()

ggplot(EDA, aes(x = fetch_projected, y = presence_absence_typ)) +
  geom_point(color = "red", size = 3, alpha = 0.7) +
  labs(title = "Scatterplot of Presence/Absence (TYP) Against Fetch", 
       x = "Fetch", 
       y = "Presence/Absence") +
  theme_minimal()
```

#### Density Against Fetch

```{r}
#| echo: false 

ggplot(EDA, aes(x = fetch_projected, y = density_sch)) +
  geom_point(color = "red", size = 3, alpha = 0.7) +
  labs(title = "Scatterplot of Density (SCH) Against Fetch", 
       x = "Fetch", 
       y = "Density") +
  theme_minimal()

ggplot(EDA, aes(x = fetch_projected, y = density_typ)) +
  geom_point(color = "red", size = 3, alpha = 0.7) +
  labs(title = "Scatterplot of Density (TYP) Against Fetch", 
       x = "Fetch", 
       y = "Density") +
  theme_minimal()
```

### Water Depth

#### Presence/Absence Against Water Depth

```{r}
ggplot(EDA, aes(x = water_depth, y = presence_absence_sch)) +
  geom_point(color = "red", size = 3, alpha = 0.7) +
  labs(title = "Scatterplot of Presence/Absence (SCH) Against Water Depth", 
       x = "Water Depth", 
       y = "Presence/Absence") +
  theme_minimal()

ggplot(EDA, aes(x = water_depth, y = presence_absence_typ)) +
  geom_point(color = "red", size = 3, alpha = 0.7) +
  labs(title = "Scatterplot of Presence/Absence (TYP) Against Water Depth", 
       x = "Water Depth", 
       y = "Presence/Absence") +
  theme_minimal()
```

#### Density Against Water Depth

```{r}
#| echo: false 

ggplot(EDA, aes(x = water_depth, y = density_sch)) +
  geom_point(color = "red", size = 3, alpha = 0.7) +
  labs(title = "Scatterplot of Density (SCH) Against Water Depth", 
       x = "Water Depth", 
       y = "Density") +
  theme_minimal()

ggplot(EDA, aes(x = water_depth, y = density_typ)) +
  geom_point(color = "red", size = 3, alpha = 0.7) +
  labs(title = "Scatterplot of Density (TYP) Against Water Depth", 
       x = "Water Depth", 
       y = "Density") +
  theme_minimal()
```

### Slope

#### Presence/Absence Against Slope

```{r}
#| echo: false 

ggplot(EDA, aes(x = slope_use, y = presence_absence_sch)) +
  geom_point(color = "red", size = 3, alpha = 0.7) +
  labs(title = "Scatterplot of Presence/Absence (SCH) Against Slope", 
       x = "Slope", 
       y = "Presence/Absence") +
  theme_minimal()

ggplot(EDA, aes(x = slope_use, y = presence_absence_typ)) +
  geom_point(color = "red", size = 3, alpha = 0.7) +
  labs(title = "Scatterplot of Presence/Absence (TYP) Against Slope", 
       x = "Slope", 
       y = "Presence/Absence") +
  theme_minimal()
```

#### Density Against Slope

```{r}
#| echo: false 

ggplot(EDA, aes(x = slope_use, y = density_sch)) +
  geom_point(color = "red", size = 3, alpha = 0.7) +
  labs(title = "Scatterplot of Density (SCH) Against Slope", 
       x = "Slope", 
       y = "Density") +
  theme_minimal()

ggplot(EDA, aes(x = slope_use, y = density_typ)) +
  geom_point(color = "red", size = 3, alpha = 0.7) +
  labs(title = "Scatterplot of Density (TYP) Against Slope", 
       x = "Slope", 
       y = "Density") +
  theme_minimal()
```

### Speed Limit

#### Presence/Absence Against Speed Limit (Bar Plot)

```{r}
#| echo: false 

ggplot(EDA, aes(x = factor(speed_limit), fill = factor(presence_absence_sch))) +
  geom_bar(position = "dodge") +
  labs(title = "Frequency of Presence/Absence (SCH) by Speed Limit",
       x = "Speed Limit",
       y = "Frequency",
       fill = "Presence/Absence") +
  theme_minimal()

ggplot(EDA, aes(x = factor(speed_limit), fill = factor(presence_absence_typ))) +
  geom_bar(position = "dodge") +
  labs(title = "Frequency of Presence/Absence (TYP) by Speed Limit",
       x = "Speed Limit",
       y = "Frequency",
       fill = "Presence/Absence") +
  theme_minimal()


```

#### Density Against Speed Limit

```{r}
#| echo: false 

ggplot(EDA, aes(x = speed_limit, y = density_sch)) +
  geom_point(color = "red", size = 3, alpha = 0.7) +
  labs(title = "Scatterplot of Density (SCH) Against Speed Limit", 
       x = "Speed Limit", 
       y = "Density") +
  theme_minimal()

ggplot(EDA, aes(x = speed_limit, y = density_typ)) +
  geom_point(color = "red", size = 3, alpha = 0.7) +
  labs(title = "Scatterplot of Density (TYP) Against Speed Limit", 
       x = "Speed Limit", 
       y = "Density") +
  theme_minimal()
```
